Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-14 article.tex
more examples for parse1 writeup, fix up code, tidy up explanation

@@

Comment      :=  ';'  (not '\n')(*)
\end{verbatim}
With the additional rule that whitespace and comments may appear in any 
amount before any token.  Tokens are:  {, }, (, ), and Symbol.

First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
can be built out of the StateT monad transformer applied to Maybe, so the new
parser type is:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}
And now we can build the item parser using the MonadState combinators:
\begin{verbatim}

\end{verbatim}

Along the way, we'll also explore some of the consequences of combining monads,
both with respect to semantics and to the definition of the Haskell language.  We'll
also bump in to some of the limitations, trade-offs and design decisions of the 
transformer/mtl library.

@@

The first parser we build will be responsible for determining whether input
text conforms to the language definition and simultaneously build an 
Abstract Syntax Tree representing of the structure of the recognized input.

Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ANumber Integer
    | ASymbol String
    | AString String
    | ALambda [String] [AST]
    | ADefine String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}
Plus we'll need a type for our parsers.  We'll use the basic type mentioned 
above -- \begin{verbatim}[t] -> m (a, [t]) \end{verbatim}
 -- as our starting point.  This type corresponds to our first monad transformer:  
 StateT (from Control.Monad.State), so we can write:
\begin{verbatim}
type Parser a = StateT [Char] Maybe a
\end{verbatim}
and a function for running a parser:
\begin{verbatim}
runParser :: Parser a -> String -> Maybe (a, String)
runParser = runStateT
\end{verbatim}
Which means that our Parser is a State monad transformer, with the Maybe type
as its underlying Monad parameter.

The significance of the Maybe parameter is two-fold:  it allows computational
effects to express first success versus failure, using the Just and Nothing 
constructors, respectively, and second backtracking/alternation.  We'll
capture both of these characteristics with using a type class, Plus:
\begin{verbatim}
class Applicative f => Plus f where
  zero :: f a
  (<+>) :: f a -> f a -> f a
\end{verbatim}
This type class is identical in intent to the Alternative type class from
Control.Applicative, with zero corresponding to empty and <+> to \verb+<|>+; 
however, we aren't able to use Alternative directly because
other of the monad transformers have unsuitable instances for our purposes,
and as we'll see later, inability to swap out such instances is one of the 
limitations of the monad transformer approach.

To get around this limitation, we'll create instances for the Plus type class:
\begin{verbatim}
instance Plus Maybe where
  zero = Nothing
  Nothing <+> y = y
  x       <+> _ = x

instance (Monad m, Plus m) => Plus (StateT s m) where
  zero = StateT (const zero)
  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
\end{verbatim}
Note that the StateT instance relies on its monad parameter, m, also being a 
member of the Plus type class.  This works out for our parser, since m is Maybe
and we do have a Plus instance for Maybe.

\subsection{Token parsers}
Now let's get down to the business of building parsers, starting with parsers
for our most basic syntactic elements:  tokens.
These include the four braces, symbols, strings, numbers, whitespace, and comments.
To recognize braces, we need to re-implement the item and check combinators that
we did before.  Using the MonadState and Plus type classes, 
we can now implement the basic item parser as a stateful
computation that fails if the state is empty:
\begin{verbatim}
item :: (MonadState [t] m, Plus m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> zero;
\end{verbatim}
And we can easily implement a new version of the check combinator:
\begin{verbatim}
check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
check f p =
    p >>= \x ->
    if (f x) then return x else zero
\end{verbatim}
And use both of them to build a new combinator, satisfy, that checks whether
the next character meets a condition:
\begin{verbatim}
satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
satisfy = flip check item
\end{verbatim}
And finally a parser that checks that the next character is equal to a
given character:
\begin{verbatim}
literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
literal c = satisfy ((==) c)
\end{verbatim}
Now we're ready to build parsers for the four bracket tokens.  
An exercise for the reader:  what is the inferred type of each parser?
Let's first build a parser for open curly brackets:
\begin{verbatim}
ocurly = literal '{'
\end{verbatim}
And let's try it out in ghci to make sure it does what we expect:
\begin{verbatim}
ghci$ runParser ocurly "{abc"
Just ('{', "abc")

ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
[Nothing, Nothing, Nothing]
\end{verbatim}
Good, it correctly accepts open curly brackets and rejects everything else.
But what about the spec that said that comments and whitespace can occur 
before any token -- will our parser be able to handle that?
\begin{verbatim}
ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
Nothing
\end{verbatim}
Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a combinator (which we'll call munch) that takes in a parser, 
throws away all leading junk, then runs the parser:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')

munch p = many0 (whitespace <+> comment) *> p
\end{verbatim}
The comment parser uses a new combinator, not1, which is built out of a new type class,
Switch, whose intention is to model computations which can be switched from successful
to failing and vice versa:
\begin{verbatim}
class Switch f where
  switch :: f a -> f ()

@@

not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
not1 p = switch p *> item
\end{verbatim}
And now let's fix the ocurly parser using munch and try it out:
\begin{verbatim}
ocurly = munch $ literal '{'

ghci$ map (runParser ocurly) [";abc\n  {def", "{def", "(def", "}def"]
[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
\end{verbatim}
Excellent!  Now the open curly parser correctly throws away whitespace and comments,
and succeeds, consuming one character, when the first non-junk character is an open 
curly brace, and fails otherwise.
We can now define parsers for the other three brace tokens:
\begin{verbatim}
ccurly = munch $ literal '}'
oparen = munch $ literal '('
cparen = munch $ literal ')'
\end{verbatim}
The parser for our last token, Symbol, is slightly more complicated because it allows
multiple characters and multiple lengths.  We'll use the satisfy combinator along with 
a function that checks whether a character is one of the legal symbol characters:
\begin{verbatim}
symbol = munch $ many1 char
  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
\end{verbatim}
Note the use of the many1 combinator to introduce repetition of any length of at least one,
and we're again using the munch combinator to throw away leading junk.

\subsection{Syntactic structures}
Whereas our token parsers dealt with the syntactic primitives of the Woof grammar, 
the remaining parsers will have to implement grammar rules that combine the tokens
into syntactic structures.
The first combining form is a function application.  The rule says that it's delimited
by matching parentheses, in between which must appear one form as the operator, followed
by any number of forms as the arguments to which the operator is applied.  We can
say that like so:
\begin{verbatim}
application =
    oparen      >>
    form        >>= \op ->
    many0 form  >>= \args ->
    cparen      >>
    return (AApp op args)
\end{verbatim}
Although this implementation uses the monadic operators >> and >>=, it could
equivalently have been implemented using the applicative combinators *> and <*>:
\begin{verbatim}
application' =
    oparen        *>
    pure AApp    <*>
    form         <*>
    many0 form   <*
    cparen
\end{verbatim}
Next, we move on to define, which is also a straightforward translation from the grammar:
\begin{verbatim}
define =
    ocurly                       *>
    check (== "define") symbol   *>
    pure ADefine                <*>
    symbol                      <*>
    form                        <*
    ccurly
\end{verbatim}
Lambda is more complicated -- not only does it have additional syntax with its parameter list,
but we also need to ensure that the parameter names are unique.  We can do that using the 
check combinator again:
\begin{verbatim}
lambda =
    ocurly                          >>
    check (== "lambda") symbol      >>
    ocurly                          >>
    check distinct (many0 symbol)   >>= \params ->
    ccurly                          >>
    many1 form                      >>= \bodies ->
    ccurly                          >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed, and parsing will
continue; if there's a repeated symbol, the parsing will stop with a failure.

\subsection{Finishing touches}
Now all we have left to do is to define the special, form (which we've already used to build
the previous parsers), and woof parsers.  The special parser is simple:  it parses either 
define or lambda:
\begin{verbatim}
special = define <+> lambda
\end{verbatim}
The form parser parsers either a symbol, application, or a special form.  We could just write:
\begin{verbatim}
form = symbol <+> application <+> special
\end{verbatim}
except that the types don't match -- symbol's type parameter is a String, whereas the other two
have ASTs, so we can just fix that by mapping the correct function over symbol:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}
The final parser, woof, needs to not only parse all the forms, but also make sure that the
entire input is consumed.  The end of input check is done with the end parser:
\begin{verbatim}
end = switch item
\end{verbatim}
And we also need to munch any trailing comments and whitespace, so our final woof parser is:
\begin{verbatim}
woof = many1 form <* munch end
\end{verbatim}

\subsection{Examples}
Let's look at a few examples of our parser in action:

@@

\subsection{Error messages}
Now we get to the meat of this section -- modifying the parsers to report accurate and useful errors.
Let's start out with a list, in code, of all the errors that we're going to deal with:
\begin{verbatim}
eaOp   =  "application: missing operator"
eaCls  =  "application: missing close parenthesis"
edSym  =  "define: missing symbol"
edForm =  "define: missing form"
edCls  =  "define: missing close curly"
elPL   =  "lambda: missing parameter list"
elPrms =  "lambda: duplicate parameter names"
elPCls =  "lambda: missing parameter list close curly"
elBody =  "lambda: missing body form"
elCls  =  "lambda: missing close curly"
esName =  "special form: unable to parse"
ewUnp  =  "woof: unparsed input"
\end{verbatim}
The MonadError type class provides two useful combinators, throwError for generating errors, and 
catchError for dealing with them.  We can ignore catchError because we don't want to try to recover
from errors, we just want them reported.  Now the question is, when should an error be created
and what data should it include?  \verb+throwError+ is too basic for our needs, so we'll introduce
the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Plus m) => e -> m a -> m a
commit err p = p <+> throwError err
\end{verbatim}
This combinator takes two arguments: 1. an error value, and 2. a monadic computation.  It first tries
to run the computation, and if the computation fails, generates an error whose contents are the first
argument.  The definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+.
So what data do we want to include in our errors?  A String describing the error and a token reporting
the position should describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}
Now lets start with \verb+Application+ as our first modified, error-reporting parser:
\begin{verbatim}
application =
    oparen                       >>= \open ->
    commit (eaOp, open) form     >>= \op ->
    many0 form                   >>= \args ->
    commit (eaCls, open) cparen  >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value in \verb+open+, and continuing
with the parsing; if either \verb+form+ or close-parenthesis fails, an appropriate error can be generated
which includes a string and the location of the open parenthesis.  However, if the open parenthesis is not
successfully parsed, it's a failure but not an error.  Here are some examples to demonstrate this in action:
\begin{verbatim}
ghci$ runParser application (countLineCol "oops")
Right' Nothing

ghci$ runParser application (countLineCol "(a b c)")
Right' (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))

ghci$ runParser application (countLineCol "\n()")
Left' ("application: missing operator",('(',2,1))

ghci$ runParser application (countLineCol "(a b")
Left' ("application: missing close parenthesis",('(',1,1))
\end{verbatim}
Although the previous version of the \verb+application+ parser could be implemented with just applicative combinators, 
this version certainly can't:  \verb+open+ is used to build the following parsers, making this decidedly monadic.

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
Right' (Just ([ADefine "fgh" 
                       (ALambda ["f","x","y"] 
                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
                                 AApp (ASymbol "f") [ASymbol "y",
                                                     ASymbol "a",
                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
              []))
\end{verbatim}
Here are examples showing errors detected by the parser:
\begin{verbatim}
ghci$ map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
                                     "{define a}", "{define a b", 
                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
Left ("application: missing operator",('(',1,1))
Left ("application: missing close parenthesis",('(',1,1))
Left ("define: missing symbol",('{',1,1))
Left ("define: missing form",('{',1,1))
Left ("define: missing close curly",('{',1,1))
Left ("lambda: missing parameter list",('{',1,1))
Left ("lambda: duplicate parameter names",('{',1,1))
Left ("lambda: missing parameter list close curly",('{',1,1))
Left ("lambda: missing body form",('{',1,1))
Left ("lambda: missing close curly",('{',1,1))
Left ("special form: unable to parse",('{',1,1))
Left ("woof: unparsed input",(',',1,2))
\end{verbatim}

>>>>

First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
can be built out of the StateT monad transformer applied to Maybe, so the new
parser type is:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}
Now we can build the item parser using the combinators from MonadState, since
its token lists are the 'state' that it modifies.
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

import Control.Monad.State       (MonadState (..), StateT(..))

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that we needed to use the FlexibleContexts language extension for the 
MonadState constraints in the type signatures.  This formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.

@@

Comment      :=  ';'  (not '\n')(*)
\end{verbatim}

With the additional rule that whitespace and comments may appear in any 
amount before any token.  Tokens are:
\begin{itemize}
\item \verb+{+
\item \verb+}+
\item \verb+(+
\item \verb+)+
\item \verb+Symbol+
\end{itemize}

@@

\subsection{Preliminaries}
The first Woof parser we build will be responsible for determining whether input
text conforms to the language definition and simultaneously building an 
Abstract Syntax Tree representing of the structure of the recognized input.
Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ANumber Integer
    | ASymbol String
    | AString String
    | ALambda [String] [AST]
    | ADefine String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}

We'll also want a function for running our parsers, using the type given in the
prevous section:
\begin{verbatim}
runParser :: Parser [t] a -> [t] -> Maybe (a, [t])
runParser = runStateT
\end{verbatim}

\subsection{Token parsers}
Now let's get down to the business of building parsers, starting with parsers
for our most basic syntactic elements:  tokens.
These include the four braces, symbols, strings, numbers, whitespace, and comments.
To recognize braces, we need to use the item and check combinators to build a new
 combinator, satisfy, that checks whether the next character meets a condition:
\begin{verbatim}
satisfy :: (MonadState [t] m, Alternative m) => (t -> Bool) -> m t
satisfy = flip check item
\end{verbatim}

And a combinator that checks that the next character is equal to a given character:
\begin{verbatim}
literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
literal c = satisfy ((==) c)
\end{verbatim}

Now we're ready to build parsers for the four bracket tokens.  
First, a simple bracket parser:
\begin{verbatim}
opencurly = literal '{'
\end{verbatim}

And some examples to make sure it's behaving correctly:
\begin{verbatim}
*Main> map (runParser  opencurly) ["{abc", "}abc", "(abc", "abc"]
[Just ('{', "abc"), Nothing, Nothing, Nothing]
\end{verbatim}

Good, it correctly accepts open curly brackets and rejects everything else.
But what about the spec that said that comments and whitespace can occur 
before any token -- will our parser be able to handle that?
\begin{verbatim}
*Main> map (runParser  opencurly) ["}abc", "(abc", "abc"]
Nothing
\end{verbatim}

Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a \verb+munch+ combinator that takes in a parser, 
throws away all leading junk, then runs the parser:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')

munch p = many0 (whitespace <+> comment) *> p
\end{verbatim}
Here are these three parsers in action:
\begin{verbatim}
*Main> runParser whitespace "\n \t   \tabc"
Just ("\n \t   \t","abc")

*Main> runParser whitespace "abc"
Nothing

*Main> runParser comment ";123  \t\nabc"
Just (";123  \t","\nabc")

*Main> runParser (munch $ literal 'a') "abc"
Just ('a',"bc")

*Main> runParser (munch $ literal 'a') ";comment!\n   abc"
Just ('a',"bc")

*Main> runParser (munch $ literal 'a') ";comment!\n   bca"
Nothing

*Main> runParser (munch $ literal 'a') "bca"
Nothing
\end{verbatim}

The comment parser uses a new combinator, not1, which is built out of a new
type class, Switch, which models computations which can be switched from
successful to failing and vice versa:
\begin{verbatim}
class Switch f where
  switch :: f a -> f ()

@@

not1 :: (MonadState [t] m, Alternative m, Switch m) => m a -> m t
not1 p = switch p *> item
\end{verbatim}

With munch in hand, we can now fix opencurly and try it out:
\begin{verbatim}
opencurly = munch $ literal '{'

*Main> map (runParser opencurly) [";abc\n  {def", "{def", "(def", "}def"]
[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
\end{verbatim}
Excellent!  Now the open curly parser correctly throws away whitespace and comments,
and succeeds, consuming one character, when the first non-junk character is an open 
curly brace, and fails otherwise.
It's straightforward to define parsers for the other three brace tokens:
\begin{verbatim}
closecurly = munch $ literal '}'
openparen  = munch $ literal '('
closeparen = munch $ literal ')'
\end{verbatim}

The parser for our last token, Symbol, is slightly more complicated because it allows
multiple characters and multiple lengths.  We'll use the satisfy combinator along with 
a function that checks whether a character is one of the legal symbol characters:
\begin{verbatim}
symbol = munch $ many1 char
  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))

*Main> runParser symbol "abc123"
Just ("abc","123")

*Main> runParser symbol ";sdfasdfsa\n  abc123"
Just ("abc","123")

*Main> runParser symbol ";sdfasdfsa\n  123abc"
Nothing
\end{verbatim}
Note the use of many1 to introduce repetition,
and we're again using munch to throw away leading junk.

\subsection{Syntactic structures}
Whereas our token parsers dealt with the syntactic primitives of the Woof grammar, 
the remaining parsers will have to implement grammar rules that combine the tokens
into syntactic structures.
The first combining form is a function application.  The rule says that it's delimited
by matching parentheses, in between which must appear one form as the operator, followed
by any number of forms as the arguments to which the operator is applied.  We can
say that like so:
\begin{verbatim}
application =
    openparen    *>
    pure AApp   <*>
    form        <*>
    many0 form  <*
    closeparen
\end{verbatim}
Note that, for now, we're doing all parser sequencing using the Applicative
combinators -- we don't need to use the monadic ones.  Also, we used pure to
inject a value -- the function AApp -- into the parser.  pure has no effect on
the tokens and always succeeds:
\begin{verbatim}
*Main> runParser (pure 3) ""
Just (3,"")

*Main> runParser (pure 3) "abc"
Just (3,"abc")
\end{verbatim}

Next, we move on to define, which is also a straightforward translation from the grammar:
\begin{verbatim}
define =
    opencurly                    *>
    check (== "define") symbol   *>
    pure ADefine                <*>
    symbol                      <*>
    form                        <*
    closecurly
\end{verbatim}

Lambda is more complicated -- not only does it have additional syntax with its parameter list,
but we also need to ensure that the parameter names are unique.  We can do that using the 
check combinator again:
\begin{verbatim}
lambda = 
    opencurly                       *>
    check (== "lambda") symbol      *>
    opencurly                       *>
    pure ALambda                   <*>
    check distinct (many0 symbol)  <*>
    (closecurly                     *>
     many1 form                    <*
     closecurly)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed, and parsing will
continue; if there's a repeated symbol, the parsing will stop with a failure.

\subsection{Finishing touches}
Now all we have left to do is to define the special, form (which we've already used to build
the previous parsers), and woof parsers.  The special parser is simple:  it parses either 
define or lambda:
\begin{verbatim}
special = define <+> lambda
\end{verbatim}

A form is either a symbol, application, or a special form.  We could just write:
\begin{verbatim}
form = symbol <+> application <+> special
\end{verbatim}

except that the types don't match -- symbol's type parameter is a String, whereas the other two
have ASTs, so we fix that by mapping the ASymbol function over symbol:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}

The final parser, woof, needs to not only parse all the forms, but also make sure that the
entire input is consumed.  The end of input check is done with the end parser:
\begin{verbatim}
end = switch item
\end{verbatim}

And we also need to munch any trailing comments and whitespace, so our final woof parser is:
\begin{verbatim}
woof = many1 form <* munch end
\end{verbatim}

\subsection{Examples}
Let's look at a few examples of our parser in action:
\begin{verbatim}
*Main> runParser woof "; \n   {lambda {x y z} a b (c b a)}end"
Just ([ALambda ["x","y","z"] 
               [ASymbol "a",
                ASymbol "b",
                AApp (ASymbol "c") 
                     [ASymbol "b",
                      ASymbol "a"]],
       ASymbol "end"],
      "")

*Main> runParser woof "a b c {define q (f x)}"
Just ([ASymbol "a",
       ASymbol "b",
       ASymbol "c",
       ADefine "q" (AApp (ASymbol "f") [ASymbol "x"])],"")

*Main> runParser woof "a b c {define q (f x)},"
Nothing
\end{verbatim}
Note that the first two parses consume the entire input, which the last one 
fails because it can't recognize the trailing comma.

@@

\subsection{Error messages}
Now we get to the meat of this section -- modifying the parsers to report accurate and useful errors.
Let's start out with a list, in code, of all the errors that we're going to deal with:
\begin{verbatim}
eaOp   =  "application: missing operator"
eaCls  =  "application: missing close parenthesis"
edSym  =  "define: missing symbol"
edForm =  "define: missing form"
edCls  =  "define: missing close curly"
elPL   =  "lambda: missing parameter list"
elPrms =  "lambda: duplicate parameter names"
elPCls =  "lambda: missing parameter list close curly"
elBody =  "lambda: missing body form"
elCls  =  "lambda: missing close curly"
esName =  "special form: unable to parse"
ewUnp  =  "woof: unparsed input"
\end{verbatim}
The MonadError type class provides two useful combinators, throwError for generating errors, and 
catchError for dealing with them.  We can ignore catchError because we don't want to try to recover
from errors, we just want them reported.  Now the question is, when should an error be created
and what data should it include?  \verb+throwError+ is too basic for our needs, so we'll introduce
the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Plus m) => e -> m a -> m a
commit err p = p <+> throwError err
\end{verbatim}
This combinator takes two arguments: 1. an error value, and 2. a monadic computation.  It first tries
to run the computation, and if the computation fails, generates an error whose contents are the first
argument.  The definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+.
So what data do we want to include in our errors?  A String describing the error and a token reporting
the position should describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}
Now lets start with \verb+Application+ as our first modified, error-reporting parser:
\begin{verbatim}
application =
    oparen                       >>= \open ->
    commit (eaOp, open) form     >>= \op ->
    many0 form                   >>= \args ->
    commit (eaCls, open) cparen  >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value in \verb+open+, and continuing
with the parsing; if either \verb+form+ or close-parenthesis fails, an appropriate error can be generated
which includes a string and the location of the open parenthesis.  However, if the open parenthesis is not
successfully parsed, it's a failure but not an error.  Here are some examples to demonstrate this in action:
\begin{verbatim}
*Main> runParser application (countLineCol "oops")
Right' Nothing

*Main> runParser application (countLineCol "(a b c)")
Right' (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))

*Main> runParser application (countLineCol "\n()")
Left' ("application: missing operator",('(',2,1))

*Main> runParser application (countLineCol "(a b")
Left' ("application: missing close parenthesis",('(',1,1))
\end{verbatim}
Although the previous version of the \verb+application+ parser could be implemented with just applicative combinators, 
this version certainly can't:  \verb+open+ is used to build the following parsers, making this decidedly monadic.

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
Right' (Just ([ADefine "fgh" 
                       (ALambda ["f","x","y"] 
                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
                                 AApp (ASymbol "f") [ASymbol "y",
                                                     ASymbol "a",
                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
              []))
\end{verbatim}
Here are examples showing errors detected by the parser:
\begin{verbatim}
*Main> map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
                                     "{define a}", "{define a b", 
                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
Left ("application: missing operator",('(',1,1))
Left ("application: missing close parenthesis",('(',1,1))
Left ("define: missing symbol",('{',1,1))
Left ("define: missing form",('{',1,1))
Left ("define: missing close curly",('{',1,1))
Left ("lambda: missing parameter list",('{',1,1))
Left ("lambda: duplicate parameter names",('{',1,1))
Left ("lambda: missing parameter list close curly",('{',1,1))
Left ("lambda: missing body form",('{',1,1))
Left ("lambda: missing close curly",('{',1,1))
Left ("special form: unable to parse",('{',1,1))
Left ("woof: unparsed input",(',',1,2))
\end{verbatim}
